---
title: "Text data mining and analysis"
description: |
  Creating word clouds and sentiment analysis for J.K. Rowling's *Harry Potter and the Philosopher's Stone* and *Harry Potter and the Deathly Hallows*.
author:
  - name: Kirsten Hodgson
    url: {}
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidytext)
library(textdata)
library(harrypotter)
library(ggwordcloud)
library(patchwork)
```

### Project Description

Here, I wrangle text data from two books in J.K. Rowling's *Harry Potter* series into tidy format, then explore the most common non-stop words in both books, using `patchwork` to visualize my graphics side-by-side. Then, I conduct sentiment analysis (using the AFINN Lexicon) on both books to compare positivity and negativity of each book by chapter. The two books I chose were *Harry Potter and the Philosopher's Stone* and *Harry Potter and the Deathly Hallows*, made available through the `harrypotter` package on GitHub.

### 1. Wrangling the Data into Tidy Format

In this section, I download the chosen books and wrangle them into tidy format, unnesting the text into chapters and removing extra white space. Then, I create a numbered chapter column and get tokens before removing stop words using `anti_join()`. Lastly, I get counts of each words throughout the entirety of each book.

```{r message = F, warning = F}
ps_tidy <- data.frame(philosophers_stone) %>% #Primary wrangling for philosopher's stone
  mutate(text_full = str_split(philosophers_stone, pattern = "\\n")) %>% #need two \ here to override operator
  unnest(text_full) %>% #Unnest text column
  mutate(text_full = str_trim(text_full)) #Remove whitespace before and after string

dh_tidy <- data.frame(deathly_hallows) %>% #Do the same for deathly hallows
  mutate(text_full = str_split(deathly_hallows, pattern = "\\n")) %>% #need two \ here to override operator
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full))

#Create a chapter column and fill it in with sequential numbers:
ps_df <- ps_tidy %>% 
  mutate(chapter = 1:n())

dh_df <- dh_tidy %>% 
  mutate(chapter = 1:n())

#Create tokens by unnesting text into a word column, then remove original
ps_tokens <- ps_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-philosophers_stone)

#ps_count <- ps_tokens %>% 
 # count(chapter, word)

dh_tokens <- dh_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-deathly_hallows)

#Remove stop words with `anti_join` and get counts of the nonstop words for both books
ps_nonstop_words <- ps_tokens %>% 
  anti_join(stop_words) 

ps_nonstop_counts <- ps_nonstop_words %>% count(word)

dh_nonstop_words <- dh_tokens %>% 
  anti_join(stop_words) 

dh_nonstop_counts <- dh_nonstop_words %>% count(word)
```

### 2. Word Clouds

##### Exploring the most common words in *Harry Potter and the Philosopher's Stone* and *Harry Potter and the Deathly Hallows*

Here, I find the 100 most common words in both *Harry Potter* books by arranging them in decreasing order of count, then using `slice()` to keep only the first 100. I then create a word cloud for each book, using `geom_text_wordcloud()` from the `ggwordcloud` package. Finally, I use patchwork to arrange both word clouds next to each other.

```{r, fig.align="center"}
#Find the top 100 words in deathly hallows and set the angle for 30% of them to 90 degrees
dh_top_100 <-
  dh_nonstop_counts %>% 
  arrange(-n) %>% 
  slice(1:100) %>% 
  mutate(angle = 90 * sample(c(0,1), n(), replace = TRUE, prob = c(70,30)))

#Create a word cloud for Deathly Hallows where word color and size depends on frequency
dh_cloud <- ggplot(dh_top_100, aes(label = word, angle = angle)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = "circle", eccentricity = 1) +
  scale_size_area(max_size = 10) +
  theme_void() +
  scale_color_gradient(low = "darkred", high = "red")

#Find the top 100 words in philosopher's stoen and set the angle for 30% of them to 90 degrees
ps_top_100 <-
  ps_nonstop_counts %>% 
  arrange(-n) %>% 
  slice(1:100) %>% 
  mutate(angle = 90 * sample(c(0,1), n(), replace = TRUE, prob = c(70,30)))

#Create a word cloud for PS where word color and size depends on frequency
ps_cloud <- ggplot(ps_top_100, aes(label = word, angle = angle)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = "circle", eccentricity = 1) +
  scale_size_area(max_size = 10) +
  theme_void() +
  scale_color_gradient(low = "darkred", high = "red")

#Use `patchwork` package to put word clouds into one graphic and then call it
ps_dh_cloud <- (ps_cloud | dh_cloud)

ps_dh_cloud
```
**Figure 1.** Word cloud exploration of the most common words in *Harry Potter and the Philosopher's Stone* (left) and *Harry Potter and the Deathly Hallows* (right). The most common word in both novels is, unsurprisingly, "Harry". Also common are the names of other main characters such as Hermione, Ron, and Dumbledore. Another valuable analysis would be to remove these common character names prior to counting word incidence.

### 3. Sentiment Analysis

In this section, I join the AFINN lexicon to my list of non-stop words for each *Harry Potter* book, then find counts of sentiments for each chapter in order to find the mean value of sentiments by chapter. Next, I create separate column plots for each book of mean AFINN sentiment value by chapter, coloring the negative values in dark red and the positive in gold. Lastly, I use `patchwork` to display the column plots next to each other for easy comparison.

```{r message = F, warning = F, fig.align="center"}
#Get the afinn lexicon data
afinn <- get_sentiments("afinn")

#Merge afinn lexicon with nonstop words for PS, then get counts for afinn values
ps_afinn <- ps_nonstop_words %>% 
  inner_join(get_sentiments("afinn"))

ps_afinn_counts <- ps_afinn %>% 
  count(chapter, value)

#Get mean afinn values for each chapter of the whole book, and create a sign column to identify whether they are positive or negative means
ps_afinn_means <- ps_afinn %>% 
  group_by(chapter) %>% 
  summarize(mean_afinn = mean(value)) %>% 
  mutate(sign = case_when(mean_afinn >= 0 ~ "positive",
                          mean_afinn < 0 ~ "negative"))

#Merge afinn lexicon with nonstop words for DH, then get counts for afinn values
dh_afinn <- dh_nonstop_words %>% 
  inner_join(get_sentiments("afinn"))

dh_afinn_counts <- dh_afinn %>% 
  count(chapter, value)

#Get mean afinn values for each chapter of the whole book, and create a sign column to identify whether they are positive or negative means
dh_afinn_means <- dh_afinn %>% 
  group_by(chapter) %>% 
  summarize(mean_afinn = mean(value)) %>% 
  mutate(sign = case_when(mean_afinn >= 0 ~ "positive",
                          mean_afinn < 0 ~ "negative"))

#Plot the mean afinn values for both books separately, with red colors for negative values and gold for positive
ps_afinn_plot <- ggplot(data = ps_afinn_means, 
                        aes(x = chapter, 
                            y = mean_afinn)) +
  geom_col(aes(fill = sign)) +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "goldenrod2")) +
  labs(x = "Chapter", 
       y = "Mean AFINN Lexicon Score", 
       title = "Philosopher's Stone") +
  theme(legend.position = "none")

dh_afinn_plot <- ggplot(data = dh_afinn_means, 
                        aes(x = chapter, 
                            y = mean_afinn)) +
  geom_col(aes(fill = sign))  +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "goldenrod2")) +
  labs(x = "Chapter", 
       y = "Mean AFINN Lexicon Score", 
       title = "Deathly Hallows") +
  theme(legend.position = "none")

#Use patchwork to merge the two graphics into one and then call it
ps_dh_afinn_plot <- (ps_afinn_plot | dh_afinn_plot)

ps_dh_afinn_plot
```
**Figure 2.** Sentiment analysis by chapter of *Harry Potter and the Philosopher's Stone* (left) and *Harry Potter and the Deathly Hallows* (right) using the AFINN Sentiment Lexicon to calculate mean chapter sentiment on a scale of positivity/negativity. Both books are overwhelmingly negative, though *Harry Potter and the Deathly Hallows* is more so, with a larger number of chapters having more negative mean scores. In both books, only two chapters have positive mean scores. Notably, the last chapter of *Harry Potter and the Deathly Hallows* is mean positive. This aligns with the text as this chapter is the epilogue, which ends the entire series with the line "All was well".

#### Data cited:

Data mined from the *Harry Potter* series:

- Rowling, J.K. *Harry Potter and the Philosopher’s Stone*. London: Bloomsbury, 1997. Print.

- Rowling, J.K. *Harry Potter and the Deathly Hallows*. London: Bloomsbury, 2007. Print.


Data made available by:

- Boehmke, B. (2016). harrypotter: An R Package for J.K. Rowling's Harry Potter Series. https://github.com/bradleyboehmke/harrypotter

