---
title: "Text data mining and analysis"
description: |
  Creating word clouds and sentiment analysis for J.K. Rowling's *Harry Potter and the Philosopher's Stone* and *Harry Potter and the Deathly Hallows*.
author:
  - name: Kirsten Hodgson
    url: {}
date: 02-25-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(tidytext)
library(textdata)
library(harrypotter)
library(ggwordcloud)
library(patchwork)
```

```{r message = F, warning = F}
ps_tidy <- data.frame(philosophers_stone) %>% #Primary wrangling for philosopher's stone
  mutate(text_full = str_split(philosophers_stone, pattern = "\\n")) %>% #need two \ here to override operator
  unnest(text_full) %>% #Unnest text column
  mutate(text_full = str_trim(text_full)) #Remove whitespace before and after string

dh_tidy <- data.frame(deathly_hallows) %>% #Do the same for deathly hallows
  mutate(text_full = str_split(deathly_hallows, pattern = "\\n")) %>% #need two \ here to override operator
  unnest(text_full) %>% 
  mutate(text_full = str_trim(text_full))

#Create a chapter column and fill it in with sequential numbers:
ps_df <- ps_tidy %>% 
  mutate(chapter = 1:n())

dh_df <- dh_tidy %>% 
  mutate(chapter = 1:n())

#Create tokens by unnesting text into a word column, then remove original
ps_tokens <- ps_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-philosophers_stone)

#ps_count <- ps_tokens %>% 
 # count(chapter, word)

dh_tokens <- dh_df %>% 
  unnest_tokens(word, text_full) %>% 
  dplyr::select(-deathly_hallows)

#Remove stop words with `anti_join` and get counts of the nonstop words for both books
ps_nonstop_words <- ps_tokens %>% 
  anti_join(stop_words) 

ps_nonstop_counts <- ps_nonstop_words %>% count(word)

dh_nonstop_words <- dh_tokens %>% 
  anti_join(stop_words) 

dh_nonstop_counts <- dh_nonstop_words %>% count(word)
```

### Word Clouds

##### Exploring the most common words in *Harry Potter and the Philosopher's Stone* and *Harry Potter and the Deathly Hallows*
```{r, fig.align="center"}
#Find the top 100 words in deathly hallows and set the angle for 30% of them to 90 degrees
dh_top_100 <-
  dh_nonstop_counts %>% 
  arrange(-n) %>% 
  slice(1:100) %>% 
  mutate(angle = 90 * sample(c(0,1), n(), replace = TRUE, prob = c(70,30)))

#Create a word cloud for Deathly Hallows where word color and size depends on frequency
dh_cloud <- ggplot(dh_top_100, aes(label = word, angle = angle)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = "circle", eccentricity = 1) +
  scale_size_area(max_size = 10) +
  theme_void() +
  scale_color_gradient(low = "darkred", high = "red")

#Find the top 100 words in philosopher's stoen and set the angle for 30% of them to 90 degrees
ps_top_100 <-
  ps_nonstop_counts %>% 
  arrange(-n) %>% 
  slice(1:100) %>% 
  mutate(angle = 90 * sample(c(0,1), n(), replace = TRUE, prob = c(70,30)))

#Create a word cloud for PS where word color and size depends on frequency
ps_cloud <- ggplot(ps_top_100, aes(label = word, angle = angle)) +
  geom_text_wordcloud(aes(color = n, size = n), shape = "circle", eccentricity = 1) +
  scale_size_area(max_size = 10) +
  theme_void() +
  scale_color_gradient(low = "darkred", high = "red")

#Use `patchwork` package to put word clouds into one graphic and then call it
ps_dh_cloud <- (ps_cloud | dh_cloud)

ps_dh_cloud
```
**Figure 1.** Word cloud exploration of the most common words in *Harry Potter and the Philosopher's Stone* (left) and *Harry Potter and the Deathly Hallows* (right). The most common word in both novels is, unsurprisingly, "Harry". Also common are the names of other main characters such as Hermione, Ron, and Dumbledore. Another valuable analysis would be to remove these common character names prior to counting word incidence.

### Sentiment Analysis

```{r message = F, warning = F}
#Get the afinn lexicon data
afinn <- get_sentiments("afinn")

#Merge afinn lexicon with nonstop words for PS, then get counts for afinn values
ps_afinn <- ps_nonstop_words %>% 
  inner_join(get_sentiments("afinn"))

ps_afinn_counts <- ps_afinn %>% 
  count(chapter, value)

#Get mean afinn values for each chapter of the whole book, and create a sign column to identify whether they are positive or negative means
ps_afinn_means <- ps_afinn %>% 
  group_by(chapter) %>% 
  summarize(mean_afinn = mean(value)) %>% 
  mutate(sign = case_when(mean_afinn >= 0 ~ "positive",
                          mean_afinn < 0 ~ "negative"))

#Merge afinn lexicon with nonstop words for DH, then get counts for afinn values
dh_afinn <- dh_nonstop_words %>% 
  inner_join(get_sentiments("afinn"))

dh_afinn_counts <- dh_afinn %>% 
  count(chapter, value)

#Get mean afinn values for each chapter of the whole book, and create a sign column to identify whether they are positive or negative means
dh_afinn_means <- dh_afinn %>% 
  group_by(chapter) %>% 
  summarize(mean_afinn = mean(value)) %>% 
  mutate(sign = case_when(mean_afinn >= 0 ~ "positive",
                          mean_afinn < 0 ~ "negative"))
```

```{r, fig.align = "center"}
#Plot the mean afinn values for both books separately, with red colors for negative values and gold for positive
ps_afinn_plot <- ggplot(data = ps_afinn_means, 
                        aes(x = chapter, 
                            y = mean_afinn)) +
  geom_col(aes(fill = sign)) +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "goldenrod2")) +
  labs(x = "Chapter", 
       y = "Mean AFINN Lexicon Score", 
       title = "Philosopher's Stone") +
  theme(legend.position = "none")

dh_afinn_plot <- ggplot(data = dh_afinn_means, 
                        aes(x = chapter, 
                            y = mean_afinn)) +
  geom_col(aes(fill = sign))  +
  theme_minimal() +
  scale_fill_manual(values = c("darkred", "goldenrod2")) +
  labs(x = "Chapter", 
       y = "Mean AFINN Lexicon Score", 
       title = "Deathly Hallows") +
  theme(legend.position = "none")

#Use patchwork to merge the two graphics into one and then call it
ps_dh_afinn_plot <- (ps_afinn_plot | dh_afinn_plot)

ps_dh_afinn_plot
```
**Figure 2.** Sentiment analysis by chapter of *Harry Potter and the Philosopher's Stone* (left) and *Harry Potter and the Deathly Hallows* (right) using the AFINN Sentiment Lexicon to calculate mean chapter sentiment on a scale of positivity/negativity. Both books are overwhelmingly negative, though *Harry Potter and the Deathly Hallows* is more so, with a larger number of chapters having more negative mean scores. In both books, only two chapters have positive mean scores. Notably, the last chapter of *Harry Potter and the Deathly Hallows* is mean positive. This aligns with the text as this chapter is the epilogue, which ends the entire series with the line "All was well".

#### Data cited:

Data mined from the *Harry Potter* series:

Rowling, J.K. *Harry Potter and the Philosopher’s Stone*. London: Bloomsbury, 1997. Print.

Rowling, J.K. *Harry Potter and the Deathly Hallows*. London: Bloomsbury, 2007. Print.

Data made available by:

Boehmke, B. (2016). harrypotter: An R Package for J.K. Rowling's Harry Potter Series. https://github.com/bradleyboehmke/harrypotter

